---
title: "Lab 10 Homework"
author: "Geralin Love Virata"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

## Desert Ecology
For this assignment, we are going to use a modified data set on [desert ecology](http://esapubs.org/archive/ecol/E090/118/). The data are from: S. K. Morgan Ernest, Thomas J. Valone, and James H. Brown. 2009. Long-term monitoring and experimental manipulation of a Chihuahuan Desert ecosystem near Portal, Arizona, USA. Ecology 90:1708.
```{r, comments=NA}
deserts <- read_csv(here("lab10", "data", "surveys_complete.csv"))
```

1. Use the function(s) of your choice to get an idea of its structure, including how NA's are treated. Are the data tidy?  
```{r, comments=NA}
glimpse(deserts)
```
Using the glimpse() command, it appears NA's are only represented by "NA" and not values. We can then summarize the number of NA's in the data using the summarize() command, finding that there are 7,599 NA's present.
```{r, comments=NA}
deserts %>%
  summarize(number_nas = sum(is.na(deserts)))
```
```{r, comments=NA}
deserts %>%
  naniar::miss_var_summary()
```
```{r, comments=NA}
deserts
```
The data appears tidy, following the three conventions of the tidyverse. Each variable has its own column, each observation has its own row, and each value has its own cell.


2. How many genera and species are represented in the data? What are the total number of observations? Which species is most/ least frequently sampled in the study?
```{r, comments=NA}
deserts %>%
  summarise(total_genus = n_distinct(genus, na.rm=T),
            total_species = n_distinct(species, na.rm=T),
            total_observations = n())
```
```{r, comments=NA}
deserts %>%
  count(species) %>%
  arrange(desc(n))
```
Merriami is the most frequently sampled in the study with a total of 10,596 observations.
```{r, comments=NA}
deserts %>%
  count(species) %>%
  arrange(n)
```
Clarki, scultalatus, tereticaudus, tigris, uniparens, and viridis are the least frequently sampled species, all tying with a total of only 1 observation each.

The following bar graph is a visual representation of the observation counts per species.
```{r, comments=NA}
deserts %>%
  ggplot(aes(x=species)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Samples Per Species", x="Species")
```

3. What is the proportion of taxa included in this study? Show a table and plot that reflects this count.
```{r, comments=NA}
deserts %>%
  tabyl(taxa, na.rm=T)
```

```{r, message=FALSE, comments=NA, warning=F}
deserts %>%
  ggplot(aes(x=taxa, fill=taxa)) +
  geom_bar() +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Counts Per Taxa", x="Taxa")
```

4. For the taxa included in the study, use the fill option to show the proportion of individuals sampled by `plot_type.`
```{r, comments=NA}
deserts %>%
  ggplot(aes(x=taxa, fill=plot_type)) +
  geom_bar(position = "dodge") +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Samples Per Taxa and Plot Type", x="Taxa")
```

5. What is the range of weight for each species included in the study? Remove any observations of weight that are NA so they do not show up in the plot.
```{r, comments=NA, message=F, warning=F}
deserts %>%
  group_by(species) %>%
  summarise(min_weight = min(weight, na.rm=T),
            max_weight = max(weight, na.rm=T),
            mean_weight = mean(weight, na.rm=T),
            median_weight = median(weight, na.rm=T),
            n=n())
```
The table above is a summary of weight ranges for each species. Since some species have no weights recorded at all, their minimum and maximum weights are represented by "Inf" and "-Inf". We can confirm that this is the case, by filtering for one of the species whose weight summaries are represented by "Inf" (for example: clarki) and select to view its recorded weights:
```{r, comments=NA}
deserts %>%
  filter(species=="clarki") %>%
  select(weight, species)
```
```{r, comments=NA}
deserts %>%
  ggplot(aes(x=species, y=weight)) +
  geom_boxplot(na.rm=T) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Weight Per Species", x="Species", y="Weight")
```

6. Add another layer to your answer from #4 using `geom_point` to get an idea of how many measurements were taken for each species.
```{r, comments=NA, message=F}
deserts%>%
  filter(weight!="NA")%>%
  group_by(species) %>%
  summarise(total_weight_measurements = n_distinct(record_id))
```
```{r, comments=NA, message=F}
deserts %>%
  filter(weight!="NA")%>%
  group_by(species)%>%
  summarise(total_weight_measurements = n_distinct(record_id)) %>%
  ggplot(aes(x=species, y=total_weight_measurements, color=total_weight_measurements)) +
  geom_point() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Total Weight Measurements Per Species", x="Species", y="Total Weight Measurements")
```

7. [Dipodomys merriami](https://en.wikipedia.org/wiki/Merriam's_kangaroo_rat) is the most frequently sampled animal in the study. How have the number of observations of this species changed over the years included in the study?
```{r, comments=NA}
deserts %>%
  filter(species == "merriami") %>%
  count(species, year) %>%
  select(year, n)
```
```{r, comments=NA}
deserts %>%
  filter(species == "merriami") %>%
  ggplot(aes(x=year, fill=species)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Dipodomys merriami Observed Over Time",x="Year",y="Number of Observations")
```

8. What is the relationship between `weight` and `hindfoot` length? Consider whether or not over plotting is an issue.
```{r, comments=NA}
deserts %>%
  ggplot(aes(x=weight, y=hindfoot_length)) +
  geom_jitter(na.rm=T, size=0.25) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Weight Versus Hindfood Lenght",x="Weight",y="Hindfoot Length")
```

There is no distinguishable relationship between the two variables due to over plotting.

9. Which two species have, on average, the highest weight? Once you have identified them, make a new column that is a ratio of `weight` to `hindfoot_length`. Make a plot that shows the range of this new ratio and fill by sex.
```{r, message=F, comments=NA}
deserts %>%
  group_by(species) %>%
  summarise(avg_weight = mean(weight, na.rm=T)) %>%
  arrange(desc(avg_weight))
``` 
```{r, comments=NA}
 deserts %>%
  filter(species == "albigula" | species == "spectabilis") %>%
  mutate(weight_to_hindfoot = (weight/hindfoot_length)) %>%
  ggplot(aes(x=species, y=weight_to_hindfoot, fill=sex)) +
  geom_boxplot(na.rm=T) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Weight to Hindfoot Length Ratio In Heaviest Species", x="Species", y="Ratio of Weight to Hindfoot Length")
```

10. Make one plot of your choice! Make sure to include at least two of the aesthetics options you have learned.
Below, I have compared the hindfoot lengths of males and females for each genus that had available data.
```{r, message=F, comments=NA}
deserts %>%
  filter(sex != "NA" & hindfoot_length != "NA") %>%
  group_by(genus, sex) %>%
  summarise(avg_hindfoot_length = mean(hindfoot_length),
            min_hf_length = min(hindfoot_length),
            max_hf_length = max(hindfoot_length),
            median_hf_length = median(hindfoot_length),
            n=n())
```

```{r, comments=NA}
deserts %>%
  filter(sex != "NA" & hindfoot_length != "NA") %>%
  ggplot(aes(x=genus, y=hindfoot_length, fill=sex)) +
  geom_boxplot(size=0.75, na.rm=T) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Relationship between Sex and Average Hindfoot Length By Genus", x="Genus", y="Hindfoot Length")
```

It appears, that given the data we have, males and females do not differ much in their hindfoot lengths in each genus.

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 