---
title: "Lab 11 Homework"
author: "Geralin Love Virata"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

**In this homework, you should make use of the aesthetics you have learned. It's OK to be flashy!**

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
library(RColorBrewer)
library(paletteer)
library(ggthemes)
library(wesanderson)
``` 

```{r}
library("ggsci")
library("ggplot2")
library("gridExtra")

data("diamonds")

p1 = ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.05, size = 1, span = 1) +
  theme_bw()

p2 = ggplot(subset(diamonds, carat > 2.2 & depth > 55 & depth < 70),
       aes(x = depth, fill = cut)) +
  geom_histogram(colour = "black", binwidth = 1, position = "dodge") +
  theme_bw()
```

## Resources
The idea for this assignment came from [Rebecca Barter's](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/) ggplot tutorial so if you get stuck this is a good place to have a look.  

## Gapminder
For this assignment, we are going to use the dataset [gapminder](https://cran.r-project.org/web/packages/gapminder/index.html). Gapminder includes information about economics, population, and life expectancy from countries all over the world. You will need to install it before use. This is the same data that we will use for midterm 2 so this is good practice.
```{r}
#install.packages("gapminder")
library("gapminder")
```

## Questions
The questions below are open-ended and have many possible solutions. Your approach should, where appropriate, include numerical summaries and visuals. Be creative; assume you are building an analysis that you would ultimately present to an audience of stakeholders. Feel free to try out different `geoms` if they more clearly present your results.  

**1. Use the function(s) of your choice to get an idea of the overall structure of the data frame, including its dimensions, column names, variable classes, etc. As part of this, determine how NA's are treated in the data.**  
```{r}
dim(gapminder)
```
```{r}
names(gapminder)
```
```{r}
summary(gapminder)
```
```{r}
glimpse(gapminder)
```

```{r}
naniar::miss_var_summary(gapminder)
```
```{r}
gapminder
```
After going through the data set, NA's do not appear to be represented by values. The naniar function did not find any NA's represented by "NA", so we can assume that this data is complete.
```{r}
gapminder <- janitor::clean_names(gapminder)
gapminder
```

**2. Among the interesting variables in gapminder is life expectancy. How has global life expectancy changed between 1952 and 2007?**
```{r}
gapminder %>%
  ggplot(aes(x=year, y=life_exp, color="#D16103")) +
  geom_jitter() +
  scale_color_brewer(palette="Set2") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "NA") +
  labs(title = "Life Expectancy Observed from 1952 to 2007",x="Year",y="Life Expectancy")
```

By plotting all observed life expectancies for each year, there appears to be a general trend of life expectancy increasing over time. We can also take a cleaner examination of the change in life expectancy over the years of 1952 to 2007 by observing the means of life expectancy in their respective year:


```{r}
life_exp_52_07 <- 
  gapminder %>%
  group_by(year) %>%
  summarise(life_exp_means = mean(life_exp))
life_exp_52_07
```
```{r}
life_exp_52_07 %>%
  ggplot(aes(x=year, y=life_exp_means, color=life_exp_means)) +
  geom_jitter(shape=8, size=3, color="#3C5488B2") +
  geom_line() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position="NA") +
  labs(title = "Life Expectancy Observed from 1952 to 2007",x="Year",y="Life Expectancy")
```

**3. How do the distributions of life expectancy compare for the years 1952 and 2007?**

```{r}
gapminder %>%
  filter(year == "1952" | year == "2007") %>%
  group_by(year) %>%
  summarise(min_exp = min(life_exp),
            max_exp = max(life_exp),
            median_exp = median(life_exp),
            mean_exp = mean(life_exp),
            total_observations = n())
```
```{r}
gapminder %>%
  filter(year == "1952" | year == "2007") %>%
  ggplot(aes(x=year, y=life_exp, group=year, fill="#D16103")) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "NA") +
  labs(title = "Life Expectancy Observed in 1952 and 2007",x="Year",y="Life Expectancy")
```
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

**4. Your answer above doesn't tell the whole story since life expectancy varies by region. Make a summary that shows the min, mean, and max life expectancy by continent for all years represented in the data.**
```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarise(min_exp = min(life_exp),
            max_exp = max(life_exp),
            mean_exp = mean(life_exp),
            total_observations = n())
```
```{r}
gapminder%>%
  group_by(year, continent)%>%
  ggplot(aes(x=year,y=life_exp, fill=continent))+
  geom_boxplot()+
  scale_fill_rickandmorty(palette="schwifty") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Life Expectancy by Continent beteween 1952 and 2007",
       x = "Year")
```
</div>

**5. How has life expectancy changed between 1952-2007 for each continent?**
```{r}
gapminder %>%
  group_by(continent, year) %>%
  filter(between(year, 1952, 2007)) %>%
  summarise(mean_exp = mean(life_exp)) %>%
  ggplot(aes(x=year, y=mean_exp)) +
  geom_line(color="yellow") +
  geom_point(shape=7, color="#3C5488B2") +
  facet_wrap(~continent)
```
  
**6. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer?**
```{r}
gapminder %>%
  ggplot(aes(x=log10(gdp_percap), y=life_exp, color=continent, shape=continent)) +
  geom_point() + 
  scale_color_brewer(palette="Accent") +
  labs(title = "Life Expectancy Versus Per Capita GDP",x="Per Capita GDP",y="Life Expectancy")
```

For most regions, it appears that there is a positive correlation between having more money and having a longer life expectancy. 


**7. Which countries have had the largest population growth since 1952?**
```{r}
gapminder %>%
  ggplot(aes(x=year, y=pop, color=country)) +
  geom_line() + 
  geom_smooth(method="lm", se=F, formula=y~x, size=0.5) +
  geom_point() +
  theme(legend.position="NA")
```
I used the above plot to get an overview of the changes in population since 1952 for all countries, omitting the legend because it was too long and interfered with the space for the graph. Seeing that the two countries with the largest population growth had more than 250000000 people in 1952, I filtered for populations greater than that amount to created a new plot to specifically show and identify those two countries.
```{r}
gapminder %>%
  filter(pop>250000000) %>%
  ggplot(aes(x=year, y=pop, color=country)) +
  geom_line() + 
  geom_smooth(method="lm", se=F, formula=y~x, size=0.5) +
  geom_point() +
  labs(title = "Change in Population Over Time", x = "Year", y = "Population")
```

China and India had the largest population growth since 1952.

**8. Use your results from the question above to plot population growth for the top five countries since 1952.**
```{r}
pop_change <- gapminder %>%
  group_by(country) %>%
  filter(year==2007|year==1952) %>%
  mutate(change_in_pop = pop - lag(pop, default = first(pop))) %>%
  arrange(desc(change_in_pop))
pop_change %>%
  select(country, change_in_pop)
```
```{r}
gapminder %>%
  filter(country=="China" | country=="India" | country=="United States" | country=="Indonesia" | country=="Brazil") %>%
  ggplot(aes(x=year, y=pop, color=country)) +
  geom_line() + 
  geom_smooth(method="lm", se=F, formula=y~x, size=0.5) +
  geom_point() +
  labs(title = "Change in Population Over Time", x = "Year", y = "Population")
```

**9. How does per-capita GDP growth compare between these same five countries?**
```{r}
gapminder %>%
  filter(country=="China" | country=="India" | country=="United States" | country=="Indonesia" | country=="Brazil") %>%
  ggplot(aes(x=year, y=gdp_percap, color=country)) +
  geom_line() + 
  geom_smooth(method="lm", se=F, formula=y~x, size=0.5) +
  geom_point() +
  labs(title = "Per-Capita GDP Over Time", x = "Year", y = "Per-Capita GDP")
```

**10. Make one plot of your choice that uses faceting!**
```{r}
gapminder %>%
  filter(year==2007 & continent=="Asia") %>%
  arrange(pop) %>%
  top_n(5, country)
```
```{r}
lowest_pops <- gapminder %>%
  filter(country=="West Bank and Gaza" | country=="Yemen, Rep." | country=="Taiwan" | country=="Thailand" | country=="Vietnam") %>%
  ggplot(aes(x=gdp_percap)) +
  geom_density() +
  labs(title="Per-Capita GDP Density of the Lowest Populated Countries in Asia 2007", x="Population", y="Density")
lowest_pops
```
```{r}
lowest_pops+facet_wrap(~country)+
  labs(title="GDP Per Capita Distribution of the Lowest Populated Countries in Asia 2007")
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 