---
title: "Midterm 1"
author: "Geralin Love Virata"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 12 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

This exam is due by **12:00p on Thursday, January 28**.  

## Load the tidyverse
If you plan to use any other libraries to complete this assignment then you should load them here.
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Questions
**1. (2 points) Briefly explain how R, RStudio, and GitHub work together to make work flows in data science transparent and repeatable. What is the advantage of using RMarkdown in this context?**  

RStudio is a GUI (graphical user interface) that makes using R (an open source, scripting language) easier. GitHub Desktop keeps track of all the changes saved on RStudio and, when the programmer has completed and knitted their code, the file can be pushed to GitHub origins for storing and managing the data. GitHub can also make your work public to make collaboration with other programmers easier.


**2. (2 points) What are the three types of `data structures` that we have discussed? Why are we using data frames for BIS 15L?**

We have discussed data matrices, data frames, and vectors. We use data frames because it the most common way of organizing data in R, storing data of many different classes.

In the midterm 1 folder there is a second folder called `data`. Inside the `data` folder, there is a .csv file called `ElephantsMF`. These data are from Phyllis Lee, Stirling University, and are related to Lee, P., et al. (2013), "Enduring consequences of early experiences: 40-year effects on survival and success among African elephants (Loxodonta africana)," Biology Letters, 9: 20130011. [kaggle](https://www.kaggle.com/mostafaelseidy/elephantsmf).  

**3. (2 points) Please load these data as a new object called `elephants`. Use the function(s) of your choice to get an idea of the structure of the data. Be sure to show the class of each variable.**
```{r comment=NA}
elephants <- readr::read_csv("data/ElephantsMF.csv")
```
```{r comment=NA}
elephants %>%
  glimpse()
```
```{r comment=NA}
elephants %>%
  summary()
```

**4. (2 points) Change the names of the variables to lower case and change the class of the variable `sex` to a factor.**
```{r comment=NA}
elephants <- janitor::clean_names(elephants)
```
```{r comment=NA}
elephants$sex <- as.factor(elephants$sex)
elephants
```

**5. (2 points) How many male and female elephants are represented in the data?**
```{r comment=NA, message=FALSE}
elephants %>%
  count(sex, na.rm = T)
```

**6. (2 points) What is the average age all elephants in the data?**
```{r comment=NA}
elephants %>%
  summarise(mean_age = mean(age, na.rm = T),
            total_counts = n())
```

**7. (2 points) How does the average age and height of elephants compare by sex?**
```{r comment=NA, message=FALSE}
elephants %>%
  group_by(sex) %>%
  summarise(mean_age = mean(age, na.rm = TRUE),
            mean_height = mean(height, na.rm = TRUE),
            total_count = n())
```

**8. (2 points) How does the average height of elephants compare by sex for individuals over 25 years old. Include the min and max height as well as the number of individuals in the sample as part of your analysis.**
```{r comment=NA, message=FALSE}
elephants %>%
  filter(age > 25) %>%
  group_by(sex) %>%
  summarise(mean_height = mean(height, na.rm = TRUE),
            min_height = min(height, na.rm = TRUE),
            max_height = max(height, na.rm = TRUE),
            total_count = n())
```

For the next series of questions, we will use data from a study on vertebrate community composition and impacts from defaunation in [Gabon, Africa](https://en.wikipedia.org/wiki/Gabon). One thing to notice is that the data include 24 separate transects. Each transect represents a path through different forest management areas.  

Reference: Koerner SE, Poulsen JR, Blanchard EJ, Okouyi J, Clark CJ. Vertebrate community composition and diversity declines along a defaunation gradient radiating from rural villages in Gabon. _Journal of Applied Ecology_. 2016. This paper, along with a description of the variables is included inside the midterm 1 folder.  

**9. (2 points) Load `IvindoData_DryadVersion.csv` and use the function(s) of your choice to get an idea of the overall structure. Change the variables `HuntCat` and `LandUse` to factors.**
```{r comment=NA}
vertebrate <- readr::read_csv("data/IvindoData_DryadVersion.csv")
```
```{r comment=NA}
summary(vertebrate)
```
```{r comment=NA}
glimpse(vertebrate)
```

```{r comment=NA}
vertebrate$HuntCat <- as.factor(vertebrate$HuntCat)
vertebrate$LandUse <- as.factor(vertebrate$LandUse)
is.factor(vertebrate$HuntCat)
is.factor(vertebrate$LandUse)
```


**10. (4 points) For the transects with high and moderate hunting intensity, how does the average diversity of birds and mammals compare?**
```{r comment=NA}
vertebrate %>%
  filter(HuntCat == "High" | HuntCat == "Moderate") %>%
  summarise(mean_bird_diversity = mean(Diversity_BirdSpecies, na.rm = T),
            mean_mammal_diversity = mean(Diversity_MammalSpecies, na.rm = T),
            total_count = n())
```
Mammals, on average, have a greater diversity than birds in transects with high and moderate hunting intensity.

**11. (4 points) One of the conclusions in the study is that the relative abundance of animals drops off the closer you get to a village. Let's try to reconstruct this (without the statistics). How does the relative abundance (RA) of apes, birds, elephants, monkeys, rodents, and ungulates compare between sites that are less than 5km from a village to sites that are greater than 20km from a village? The variable `Distance` measures the distance of the transect from the nearest village. Hint: try using the `across` operator.**  
```{r comment=NA}
vertebrate %>%
  filter(Distance <= 5) %>%
  summarise(across(contains("RA"), mean, na.rm = T))
```
```{r comment=NA}
vertebrate %>%
  filter(Distance >= 20) %>%
  summarise(across(contains("RA"), mean, na.rm = T))
```
The relative abundance of apes, elephants, monkeys, and ungulates increases when they are farther away from a village. However, the relative abundance of birds and rodents are greater when they are closer to the village than when they are farther away.

**12. (4 points) Based on your interest, do one exploratory analysis on the `gabon` data of your choice. This analysis needs to include a minimum of two functions in `dplyr.`**
```{r comment=NA, message=FALSE}
vertebrate %>%
  filter(LandUse != "Neither") %>%
  group_by(LandUse, HuntCat) %>%
  summarise(across(contains("RA"), mean, na.rm = T))
```
Here, I analyzed the relative abundance of apes, birds, elephants, monkeys, rodents, and ungulates in areas where land with none to high hunting intensity was used for parks and logging. It appears that the relative abundance of birds and monkeys remain significantly high compared to other animals, regardless of land use and hunting intensity. In contrast, ape and elephant abundance drop significantly with increased hunting and when land is used for logging.

