---
title: "Lab 7 Homework"
author: "Geralin Love Virata"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
``` 

## Data
**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r comment=NA}
amniota <- readr::read_csv("data/amniota.csv") %>%
  janitor::clean_names()
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).
```{r comment=NA}
amphibio <- readr::read_csv("data/amphibio.csv") %>%
  janitor::clean_names()
```

## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  

```{r comment=NA}
glimpse(amniota)
```
Using the glimpse() command, I notice already that NA's are written as "-999", "-999.0", and "-999.000" (or some containing more than 3 zeroes).

```{r comment=NA}
amniota %>%
  anyNA()
```
There are no NA's represented as "NA".

**3. Do some exploratory analysis of the `amphibio` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  

```{r comment=NA}
glimpse(amphibio)
```
Using the glimpse(), it appears NA's are represented only by "NA". 

```{r comment=NA}
amphibio %>%
  anyNA()
```


**4. How many total NA's are in each data set? Do these values make sense? Are NA's represented by values?**   

```{r comment=NA}
amniota %>%
  summarize(number_nas = sum(is.na(amniota)))
```
There are no NA's represented as "NA" in the amniota data frame. However, understanding the chances of having a large data frame with zero NA's is very low, it is likely that the missing values are rather represented in a different form. Referring back to problem 2, using the glimpse() command, we know that they are at least represented by "-999".

```{r comment=NA}
amphibio %>%
  summarize(number_nas = sum(is.na(amphibio)))
```
In the amphibio data set, there are 170,691 NA's. It makes sense for a large data set have a high number of NA's. They are not represented as values as in the amniota data set.

**5. Make any necessary replacements in the data such that all NA's appear as "NA".**   
```{r comment=NA}
amniota_tidy <- amniota %>%
  na_if("-999") %>%
  na_if("-30258.711")
```
NA's may have also been represented by the value "-30258.711" because it appeared in the female_maturity_d category multiple times, which is not a value that would make sense for this category. 

After changing all NA's represented by values to appear as "NA", we can now use the summarize() command to re-check how many NA's are actually present in the amniota data set (renamed "amniota_tidy").
```{r comment=NA}
amniota_tidy %>%
  summarize(number_nas = sum(is.na(amniota_tidy)))
```
This value of total NA's now makes sense.

```{r comment=NA}
amphibio_tidy <- amphibio %>%
  na_if("-999")
```
The above code is run just in case there are NA's represented by "-999" that were not caught in the exploratory analysis in number 3.

```{r comment=NA}
amphibio_tidy %>%
  summarize(number_nas = sum(is.na(amphibio_tidy)))
```
Since the number of NA's is still the same as in number 4, it is even safer to say that NA's were only represented as "NA". 

**6. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.**  
```{r comment=NA}
naniar::miss_var_summary(amniota_tidy)
```

**7. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amphibio` data.**
```{r comment=NA}
naniar::miss_var_summary(amphibio_tidy)
```

**8. For the `amniota` data, calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?**  

```{r comment=NA}
amniota_tidy %>%
  select(class, egg_mass_g) %>% 
  group_by(class) %>%
  naniar::miss_var_summary(order=T)
```
For mammals, the percentage of NA's does make biological sense. They do not lay eggs (except for the duck-billed platypus), so having no egg masses recorded is reasonable. As for birds and reptiles, the percentage of NA's does not make **biological** sense, because both classes lay eggs and would, thus, have egg masses to be recorded. However, large data sets are often incomplete and some values do not get recorded for various reasons, so NA's take the place of the actual unknown value. 

**9. The `amphibio` data have variables that classify species as fossorial (burrowing), terrestrial, aquatic, or arboreal. Calculate the number of NA's in each of these variables. Do you think that the authors intend us to think that there are NA's in these columns or could they represent something else? Explain.**
```{r comment=NA}
amphibio %>% 
  select(fos, ter, arb, aqu) %>% 
  summarise_all(~(sum(is.na(.))))
```
The authors may have intended for the NA's to represent that the species in question does not follow a particular lifestyle. For example, if the species is classified as fossorial, then the terrestrial, aquatic, and arboreal columns for the species' row would be left as NA. 

**10. Now that we know how NA's are represented in the `amniota` data, how would you load the data such that the values which represent NA's are automatically converted?**
```{r comment=NA}
amniota_tidy_2 <- readr::read_csv(file = "data/amniota.csv", na = c("-999", "-30258.711"))
```
```{r comment=NA}
glimpse(amniota_tidy_2)
```
Using the glimpse() command, we see that the NA's that were originally represented as "-999" or "-30258.711" were automatically converted to "NA".

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  